A fast algorithm for finding efficient decisions in multi-objective problems with black-box multiextremal functions

### 2022-01-02 2022/OM&S/paper.tex
include more references in the introduction

\usepackage{color}
\usepackage{hyperref}

@@

In the field of mathematical modeling the intelligent decision-making processes, the problems of multicriterial optimization (MCO) present a wide class of models that are rich and interesting from a theoretical point of view and very important in real applications. Multicriterial problems describe decision-making situations with multiple goals which are considered as functional criteria to be optimized.

In general case, there exist three key factors determining the complexity of analyzing these models. The crucial feature corresponds to the typical case where criteria are contradictory: increasing one criterion leads to decreasing the other. Such inconsistent behavior generates the necessity of introducing a compromise in the joint analysis of criteria and consideration of a set of compromised (partial) solutions as a complete solution of a multiobjective problem (Pareto set).
The second complexity factor is associated with the dimensionality or number of varied parameters of the problem because increasing the problem dimension leads to significant growth of computational cost.

Finally, the last factor is the multiextremality of criteria. First of all, multiextremal criteria can generate complicated Pareto sets consisting, for example, of several disconnected parts. Moreover, the combination of such factors as multiextremality and greater dimension leads to significant computational complexity of the problem because the computational cost increases exponentially when dimension grows. From this point of view the multiextremal problems are the most complicated ones in multicriterial optimization and this MCO class is the subject of interest in the given paper.

@@

Another idea applies a parameterization scheme for the criteria of the problem and reduces a MCO problem to a family of single-criterion (scalar) optimization problems being problems of mathematical programming which can be solved by corresponding well-developed optimization methods. In this scheme a non-negative coefficient (weight) is assigned to each criterion which reflects numerically an ``importance'' of the criterion and on the base of such weighted criteria a scalar function (convolution) is built and minimized. Under appropriate choice of convolution the solution of the built scalar optimization problem is a partial solution of the initial multicriterial problem. Changing weights we can obtain different Pareto points.

@@

Consideration of different approaches to solving MCO problems and a relevant references can be found, for example, in monographs and papers \cite{Zhou2011,Nedjah2015} \colorbox[rgb]{1,1,0}{[1-  Zhou, Nedjah ]}.
In this paper we consider multiextremal black-box MCO problems and propose a new scheme of solving these problems. This scheme is based on maximum convolution, dimensionality reduction by means of Peano curves, information-statistical global optimization algorithm used for the first time in MCO problems. Moreover, in the framework of the scheme the information already obtained in the course of solving scalar subproblems of convolution optimization can be utilized for solving a new convolution subproblem. Briefly, the way is as follows. Before beginning the optimization of a new convolution we have at our disposal values of criteria computed already and now can recalculate them to values of the current convolution taking into account this values as initial information for optimization algorithm. This procedure accelerates significantly the optimization process.

>>>>

%\usepackage{color}
\usepackage{hyperref}

@@

In the field of mathematical modeling the intelligent decision-making processes, the problems of multicriterial optimization (MCO) present a wide class of models that are rich and interesting from a theoretical point of view and very important in real applications \cite{Marler2009,Hillermeier2005}. Multicriterial problems describe decision-making situations with multiple goals which are considered as functional criteria to be optimized.

In general case, there exist three key factors determining the complexity of analyzing these models. The crucial feature corresponds to the typical case where criteria are contradictory: increasing one criterion leads to decreasing the other. Such inconsistent behavior generates the necessity of introducing a compromise in the joint analysis of criteria and consideration of a set of compromised (partial) solutions as a complete solution of a multiobjective problem (Pareto set), see, e.g., \cite{Collette2004,Ehrgott2005}.
The second complexity factor is associated with the dimensionality or number of varied parameters of the problem because increasing the problem dimension leads to significant growth of computational cost.

Finally, the last factor is the multiextremality of criteria \cite{Pardalos2017}. First of all, multiextremal criteria can generate complicated Pareto sets consisting, for example, of several disconnected parts. Moreover, the combination of such factors as multiextremality and greater dimension leads to significant computational complexity of the problem because the computational cost increases exponentially when dimension grows. From this point of view the multiextremal problems are the most complicated ones in multicriterial optimization and this MCO class is the subject of interest in the given paper.

@@

Another idea applies a parameterization scheme for the criteria of the problem and reduces a MCO problem to a family of single-criterion (scalar) optimization problems being problems of mathematical programming which can be solved by corresponding well-developed optimization methods \cite{Collette2004,Ehrgott2005}. In this scheme a non-negative coefficient (weight) is assigned to each criterion which reflects numerically an ``importance'' of the criterion and on the base of such weighted criteria a scalar function (convolution) is built and minimized. Under appropriate choice of convolution the solution of the built scalar optimization problem is a partial solution of the initial multicriterial problem. Changing weights we can obtain different Pareto points.

@@

Consideration of different approaches to solving MCO problems and a relevant references can be found, for example, in monographs and papers \cite{Miettinen1999,Ehrgott2005,Zhou2011,Nedjah2015,Pardalos2017}.
In this paper we consider multiextremal black-box MCO problems and propose a new scheme of solving these problems. This scheme is based on maximum convolution, dimensionality reduction by means of Peano curves, information-statistical global optimization algorithm used for the first time in MCO problems. Moreover, in the framework of the scheme the information already obtained in the course of solving scalar subproblems of convolution optimization can be utilized for solving a new convolution subproblem. Briefly, the way is as follows. Before beginning the optimization of a new convolution we have at our disposal values of criteria computed already and now can recalculate them to values of the current convolution taking into account this values as initial information for optimization algorithm. This procedure accelerates significantly the optimization process.
